Adaptive Sequential Model Selection

### 2015-07-23 paper.tex
rob added cart example

>>>>

\section{Decision trees and interaction models}
In this section we discuss another application of adaptive sequential inference, to binary decision trees.
Specifically, the classification and regression tree  (CART)  method builds a decision tree in a top down manner, finding the best available split
at each node.  When considering splitting two daughter nodes, the generic CART algorithm splits the nodes in an arbitrary order, only stopping when a node
reaches a minimum size.  We instead consider a  special version of CART  algorithm:  the   "best first"  method, which order splits by their achieved reduction in loss.
This procedure is used, for example in the R package for gradient boosting, called {\tt gbm}.
The best first tree-builder is thus a sequential selection  procedure,  to which we can apply the results in this paper. Hence we can obtain independent, sequential p-values
for each successive split in the decision  tree. We now give some details and an example.

Given data $(y_i, x_{i1}, x_{i2},\ldots x_{ip})$ for $i=1,2,\ldots N$,
we choose a splitting  variable $j_1$ and split point $t_1$, leading to a model of the form
\begin{eqnarray}
\hat y_i=\beta_0 +\beta_{11} I(x_{ij_1} \leq t_1) +\beta_{12} I(x_{ij_1} > t_1)
\end{eqnarray}
Then at the second stage, we try splitting one of the two daughter nodes, and choose the split and most reduces the residual sum of squares.
Suppose that a split of the first node is chosen, using predictor $j_2$ at split point $t_2$.  This leads to the interaction model of the form
\begin{eqnarray}
\hat y_i=\beta_0 +\beta_{11} I(x_{ij_1} \leq t_1) \cdot \beta_{31} I(x_{ij_2} \leq t_2)+   \beta_{11} I(x_{ij_1} \leq t_1) \cdot  \beta_{31} I(x_{ij_2} > t_2)  +\beta_{12} I(x_{ij_1} > t_1).\cr
\label{eqn:intmodel}
\end{eqnarray}
Note that the search for possible split points can be expressed through an exhaustive set of basis functions $I(x_{ij} \leq  t_i), (x_{ij}  > t_i)$ for each $t_i \in (x_{1j}, x_{2j}, \ldots x_{nj})$.
The procedure is somewhat similar to foreward stepwise regression, but with differences.
For example the set of candidate basis functions available at each stage are a function of what has been  previously entered into the model
Nonetheless,  the selection procedure  for the first set of $k$ splits can be expressed in polyhedral from $Ay \leq b$, and the saturated and selective inference procedures can be applied.
As before, the saturated p-values have an explicit truncated Gaussian form, while computation of the selective p-values require accept-reject sampling.

Figure \ref{fig:cart} shows an example with independent standard Gaussian features, and $n=30, p=3$.
We 
consider just the first two splits, in a null case (top row) and non-null (bottom row).
In the non-null case,  the data are generated from a  model of the form (\ref{eqn:intmodel}).

As expect, the nominal p-values are far too small and hence don't control type I error.
Both the saturated and selective p-values control type I error, with the selective p-values are somewhat more powerful in the non-null case.
Rules like the forwardStop procedure could be applied  to the selective p-values to choose a ``best'' model size.
\begin{figure}[h]
  \centering
  \includegraphics[width=.65\textwidth]{figs/cart.pdf}
  \caption{\em Cumulative distribution functions of P-values for  the first two splits in a CART decision tree. In the top row, the true model is null,
  while in the bottom row, the data are generated from a  model of the form (\ref{eqn:intmodel})}
  \label{fig:cart}
\end{figure}
